\section{(Un)Decidability and the Finite Model Property}

In Section~\ref{expressivity} we show four memory logics ($\cMLRK$,
$\cMLRKE$, $\cMLRKM$ and $\cMLRKME$) which are all more expressive
than $\bml$ but less expressive than $\hlogic$. Given that $\bml$ is
decidable and $\hlogic$ is undecidable, exploring where the
decidability line lies is an intriguing question.  The main goal of
this section is to investigate this issue, together with the related
question of whether the language is sufficiently expressive to force
infinite models.

We start by investigating $\cMLRKM$ and $\cMLRKME$. We will show
that even though they are equivalent in terms of expressive power
when we allow a shift if the signature, the satisfiability problem
for $\cMLRKM$ is decidable (actually \textsc{pspace}-complete) while
$\cMLRKME$ is already undecidable. As we will show in the proof of
Theorem~\ref{} the trick will be to use a `dirty' memory. In
$\cMLRKME$, we are restricted to the class of models where the
memory is always initialized to $\emptyset$ and we can't play this
trick anymore.  Actually $\cMLRKM$ is really standing on the
decidability line, adding a single nominal to $\cMLRKM$ push the
satisfiability problem over to undecidability.

\subsection{The Decidability of $\cMLRKM$\ldots}

We will first prove that $\bml$ and $\cMLRKM$ expressively
equivalent over the class of tree-like models.  With that result
at hand, we will then prove that $\cMLRKM$ has a tree model
property, and from that decidability and \textsc{pspace}-completeness
follows.

\tb{C: Yo termino con toda la seccion 5 el Sabado}

\tb{C: Termino de escribir la demo el Sabado}

\begin{thm}\label{prop:sat-preserv-tree}
$\K$  over the signature $\tup{\prop \cup \{\mathit{known}\},\rel}$
is equivalent to $\cMLRKM$ over the class of tree models.
\end{thm}

\begin{pf}
%
% We first define, given a formula
% $\varphi$, a new formula More formally:
%
% \begin{displaymath}\label{sharp}
% \begin{array}{rcl}
% p^\sharp & = & p \quad p \in \prop\\
% \known^\sharp & = & \top \\
% (\lnot \varphi)^\sharp & = & \lnot \varphi^\sharp \\
% (\varphi_1 \land \varphi_2)^\sharp & = & \varphi_1^\sharp \land \varphi_2^\sharp \\
% (\remember \varphi)^\sharp & = & \remember \varphi^\sharp\\
% (\ttup{r} \varphi)^\sharp & = & \ttup{r} \varphi
% \end{array}
% \end{displaymath}
We start by noticing that over tree models, we can eliminate
$\known$ at modal depth 0 from the formula $\remember \varphi$.

\begin{claim}\label{lem:replace}
Let $\varphi^\sharp$ be the result of replacing all the occurrences
of $\known$ that are in $\varphi$ at modal depth zero by $\top$, and
let $\gM$ be a tree model. Then $\model, w \models \remember
\varphi$ iff $\model, w \models \varphi^\sharp$.
\end{claim}

\begin{pfclaim}
We proceed by induction. The case for $\known$, the propositional
symbols and booleans are straightforward. We analyze the other
cases:
\begin{itemize}
 \item $\varphi = \remember \psi$. $\model, w \models \remember\remember\psi$ iff $\model, w \models \remember \psi$ iff (by inductive hypothesis) $\model, w \models \psi^\sharp$ iff $\model, w \models (\psi^\sharp)^\sharp$ iff (by inductive hypothesis) $\model, w \models \remember (\psi^\sharp)$ iff $\model, w \models (\remember\psi)^\sharp$
\item $\varphi = \ttup{r} \psi$. $\model, w \models \remember \ttup{r} \psi$ iff (by definition) $\model[w], w \models \ttup{r} \psi$ iff (by definition of $\sharp$) $\model[w], w \models (\ttup{r} \psi)^\sharp$ iff (by definition) $\model, w \models (\ttup{r} \psi)^\sharp$
\end{itemize}
\end{pfclaim}

Taking the same model equivalence used to compare $\bml$ with $\cMLRK$,
the following result can be established:

Define the following translation taking $\cMLRKM$ formulas over the signature
$\diam{\prop, \rel}$ to $\bml$-formulas over the signature
$\diam{\prop \cup \{known\}, \rel}$:
$$
\begin{array}{rcl}
\Tr(p) & = & p \quad p \in \prop\\
\Tr(\known) & = & known \\
\Tr(\lnot \varphi) & = & \lnot \Tr(\varphi) \\
\Tr(\varphi_1 \land \varphi_2) & = & \Tr(\varphi_1) \land \Tr(\varphi_2) \\
\Tr(\ttup{r} \varphi) & = & \diam{r} \Tr(\varphi) \\
\Tr(\remember \varphi) & = & \Tr(\varphi^\sharp)
\end{array}
$$

We proceed by induction on $\varphi$. The propositional and boolean
cases are trivial. The $\known$ case is also easy given the
definitions. For the diamond case. Given that $\model$ is tree-like,
the remember operator has no effect beyond modal operators, so
$\model, w \models_{\cMLRKM} \ttup{r}\psi$ iff exists $v$ such that
$wRv$ and $\model,v\models_{\cMLRKM} \psi$. By inductive hypothesis,
$\model,v\models_{\cMLRKM} \psi$ iff $\model, v \models_{\bml}
\Tr(\psi)$, and by definition $\model, w \models_{\bml} \diam{r}
\Tr(\psi)$. Finally, let's see the case for remember. By
Lemma~\ref{lem:replace}, $\model, w \models_{\cMLRKM} \remember \psi$
iff $\model, w \models_{\cMLRKM} \psi^\sharp$. By inductive hypothesis,
$\model, w \models_{\bml} \Tr(\psi^\sharp)$.
\end{pf}





%\subsection{Finite model property}

% \begin{lem}\label{lem:frontera-k}
% Let $\model$ be a model and $w \in \model$. Then $(\model
% \upharpoonright n),w \bisim_n \model, w$.
% \end{lem}
%
% \begin{pf}
% In the game $E_n(\model, \model \upharpoonright n, w, w)$ the
% winning strategy for Duplicator consists in copying Spoiler moves.
% This can always be done because Spoiler cannot go beyond $n$ steps
% from $w$.
% \end{pf}

% \begin{thm}
% Let $\model$ be a model and $w \in \model$. For every $n \geq 0$,
% $\model, w \leftrightsquigarrow_n (\model \upharpoonright n), w$.
% \end{thm}
%
% \begin{pf}
% This result follows from Theorem~\ref{thm:bisim-n-to-prune} and
% Theorem~\ref{thm:bisim-tlm-n-to-modal-n}.
% \end{pf}



\begin{pro}[Tree model property]\label{prop:tree-model-property}
For every $\cMLRKM$-model
$\model$ and $w\in\model$ there is a tree-like model $\model'$ where
$w\in\model'$ and $\model,w\bisim\model',w$.
\end{pro}

\begin{pf}
Let $\model=\diam{W,R,V,S}$.  We prove the result for the unimodal
case, the generalization to the multimodal case is straightforward.
We define the model $\model'=\diam{W',R',V',S'}$ as follows. Its
domain $W'$ consists on all finite sequences $(u_0,\dots,u_n)$ such
that $u_0=w$, $n\geq 0$ and there is a path $u_0Ru_1\dots Ru_n$ in
$\model$. Define $(u_0,\dots,u_n)R(v_0,\dots,v_m)$ to hold if
$m=n+1$, $u_i=v_1$ for $i=0,\dots,n$ and $u_nRv_m$ holds in
$\model$. The valuation $V'$ is defined by setting
$(u_0,\dots,u_n)\in V'(p)$ iff $u_n\in V(p)$. Finally,
$(u_0,\dots,u_{n-1},u_n)\in S'$ iff $u_n\in\{u_0,\dots,u_{n-1}\}$ or
$u_n\in S$.

We call $s_i$ the sequence $(v_0,\dots,v_i)$ for every
$i=0,\dots,n$. Let us see that Duplicator has a winning strategy in
the game $\EF(\model,\model',w,w)$. It is sufficient to see that in
the game
$\EF(\model[v_0,\dots,v_n],\model'[s_0,\dots,s_n],v_{n+1},s_{n+1})$,
Duplicator can always answer successfully to Spoiler's move.
\begin{itemize}
\item
If Spoiler chooses $\model[v_0,\dots,v_n]$ and some $v_{n+1}$, a
successor of $v_n$, Duplicator chooses the sequence
$s_{n+1}=s_nv_{n+1}$.

\item
If Spoiler chooses $\model'[s_0,\dots,s_n]$ and $s_{n+1}=s_nv_{n+1}$
(for some $v_{n+1}$), a successor of $s_n$, Duplicator chooses the
state $v_{n+1}$.
\end{itemize}
By definition, it is easy to see that $s_{n+1}$ and $v_{n+1}$ agree.
Observe that at this stage of the game the current ``memory" of
$\model[v_0,\dots,v_n]$ is $S \cup \{v_0,\dots,v_n\}$ and the
current ``memory" of $\model'[s_0,\dots,s_n]$ is $S' \cup
\{s_0,\dots,s_n\}$. Then it is also clear that $v_{n+1}$ was
remembered at this stage of the game, if and only if $s_{n+1}$ also
was. More formally, on the one hand, $v_{n+1}\in S \cup
\{v_0,\dots,v_n\}$ implies $s_{n+1} \in S'$ by definition. On the
other hand if $s_{n+1}\in S' \cup \{s_0,\dots,s_n\}$, we have
$s_{n+1}\in S'$ (since there are no cycles in $\model'$) and by
definition $v_{n+1}\in S \cup \{v_0,\dots,v_n\}$.
\end{pf}

%If $\model$ is a $\cMLRKM$-model, we denote with $\model_T$ the
%tree-like model constructed in the above proposition.

%\begin{lem}\label{lem:bisim_bml_then_bisim_tlm}
%Let $\model,\nodel$ be two $\tlm$ tree-like models. For any $n$, if
%the equivalent $\bml$-models satisfy
%$\model,w\bisim_n^{\bml}\nodel,v$ then
%$\model,w\bisim_n^{\tlm}\nodel,v$.
%\end{lem}
%
%\begin{pf}
%Let $\model=\diam{W_1,R_1,V_1,S_1}$ and
%$\nodel=\diam{W_2,R_2,V_2,S_2}$.
%%
%Recall the definition of $n$-bisimulation for $\bml$
%\cite[Definition 2.30]{BRV01}, involving the relations
%$Z_0\supseteq\dots\supseteq Z_{n}$. We define a winning strategy for
%Duplicator in terms of $Z_0,\dots,Z_{n}$.
%
%We start in the game $E(\model,\nodel,w,v)$. We show that if
%$E(\model',\nodel',w',v')$ is some $i$th stage of that game ($i\leq
%n$) and $w'Z_{n-i}v'$, then for any choice $u$ of Spoiler,
%Duplicator can always answer with $t$ such that $uZ_{n-i-1}t$.
%Suppose Spoiler chooses $u\in\model'$ such that $w'Ru$. By
%definition of $Z_{n-i}$, there exists $t\in\nodel'$ such that $v'Rt$
%and $vZ_{n-i-1}t$. It follows from the definition of $Z_0$ that $u$
%and $t$ agree. Observe that the game $E(\model',\nodel',w',v')$ is
%just equivalent to $E(\model,\nodel,w',v')$, since both $\model$ and
%$\nodel$ are acyclic. Therefore $u\in S_1$ iff $t\in S_2$. The case
%Spoiler chooses to play in $\nodel'$ is similar and this completes
%the proof.
%\end{pf}

%\begin{thm}\label{thm:fmp-tlm}
%Let $\model$ be a model. For every $n$ there is a finite tree-like
%model $\nodel$ such that $\model\bisim_n\nodel$.
%\end{thm}

%\begin{pf}
%By Proposition~\ref{prop:tree-model-property},
%$\model\bisim^{\tlm}\model_T$. By
%Theorem~\ref{thm:bisim-n-to-prune},
%$\model_T\bisim_n^{\tlm}\model_T\upharpoonright n$. Applying the
%selection process in \cite[Theorem 2.34]{BRV01} on the equivalent
%$\bml$-model for $\model_T\upharpoonright n$ we obtain  a finite
%tree-like model $\nodel$ such that $\model_T\upharpoonright
%n\bisim_n^{\bml}\nodel$. Applying
%Lemma~\ref{lem:bisim_bml_then_bisim_tlm}, we know that
%$\model_T\upharpoonright n\bisim_n^{\tlm}\nodel$ and therefore
%$\model \bisim_n^{\tlm}\nodel$.
%\end{pf}

\begin{thm}\label{thm:fmp}
If $\varphi\in\cMLRKM$ is satisfiable then it is satisfiable on a
recursively bounded model.
\end{thm}

\begin{pf}
Let $\varphi$ be a $\cMLRKM$ formula of degree $k$, and suppose
$\model,w \models_{\cMLRKM} \varphi$. By
Proposition~\ref{prop:tree-model-property}, there is a tree-like
model $\model'$ such that $\model',w \models_{\cMLRKM} \varphi$.
Using Proposition~\ref{prop:sat-preserv-tree}, we can turn to basic
modal logic $\bml$, and assert that $\model',w \models_{\bml}
\Tr(\varphi)$. Now we can use the finite tree model property for
basic modal logic~\cite{BRV01}, so there must be a finite tree-like
model $\model''$ such that $\model'',w \models_{\bml} \Tr(\varphi)$.
Finally, we can use Proposition~\ref{prop:sat-preserv-tree} again,
and conclude $\model'',w \models_{\cMLRKM} \varphi$.

Observe that it is known~\cite{BRV01} that $\model''$ has at most
$n^k$ states, where $n$ is the number of non-equivalent formulas of
degree at most $k$, containing the same propositional symbols of
$\varphi$.
\end{pf}


%\begin{cor}[Finite model property]
%If $\varphi\in\tlm$ is satisfiable then it is satisfiable on a
%finite model.
%\end{cor}

%\begin{thm}[Alternative proof for the finite model property] (without using Lemma~\ref{lem:bisim_bml_then_bisim_tlm})
%If $\varphi\in\tlm$ is satisfiable then it is satisfiable on a
%finite model.
%\end{thm}
%\begin{pf}
%Let $\varphi$ be a $\tlm$ formula, and suppose $\model,w \models
%\varphi$. By Proposition~\ref{prop:tree-model-property}, there is a
%tree-like model $\model'$ such that $\model',w \models \varphi$.
%Using Proposition~\ref{prop:sat-preserv-tree}, we can turn to basic
%modal logic $\bml$, and assert that $\model',w \models_{\bml}
%Tr(\varphi)$. Now we can use the finite tree model property for
%basic modal logic \cite{??}, so there must be a finite tree-like
%model $\model''$ such that $\model'',w \models_{\bml} Tr(\varphi)$.
%Finally, we can use Proposition~\ref{prop:sat-preserv-tree} again,
%and conclude $\model'',w \models \varphi$.
%\end{pf}


\begin{cor}[Decidability]\label{cor:tlm-decidability}
$\cMLRKM$ is decidable.
\end{cor}

\begin{pf}
By Theorem~\ref{thm:fmp}, given a $\cMLRKM$-formula $\varphi$ one
can recursively obtain a bound $b$ such that if $\varphi$ is
satisfiable then it is satisfiable in a model of at most $b$ states.
Hence, to decide the satisfiability of $\varphi$ one can check if
$\model\models\varphi$ for all models $\model$ with size at most
$b$.
\end{pf}

\begin{thm}[\textsc{pspace}-complete]
The satisfaction problem for $\cMLRKM$ is \textsc{pspace}-complete.
\end{thm}
\begin{pf}
The lower \textsc{pspace}-hard bound for $\cMLRKM$ follows from the
lower bound for the basic modal logic~\cite{BRV01}. We show a
matching upper bound by reducing the satisfiability problem for
basic modal logic. Given a $\cMLRKM$-formula $\varphi$, we compute
$\Tr(\varphi)$ using Definition~\ref{def:tr-tlm-k}. Note that the
size of the translated formula $\Tr(\varphi)$ is actually at most
the size of $\varphi$. Let suppose that $\Tr(\varphi)$ has a model,
that is $\model, w \models_{\bml} \Tr(\varphi)$. This happens iff
there exists a tree-like model $\model'$ such that $\model', w
\models_{\bml} \Tr(\varphi)$ (by the tree model property of $\bml$)
iff $\model', w \models_{\cMLRKM} \varphi$ (by
Proposition~\ref{prop:sat-preserv-tree}) iff there exists a model
$\nodel$ such that $\nodel,w \models_{\cMLRKM} \varphi$ (by the tree
model property for $\cMLRKM$, stated in
Proposition~\ref{prop:tree-model-property}). Therefore, given
$\varphi$, determining the $\bml$-satisfiability of $\Tr(\varphi)$
(and we can do this with a \textsc{pspace} algorithm~\cite{BRV01})
is enough to decide the $\cMLRKM$-satisfiability of $\varphi$.
\end{pf}

\subsection{\ldots and its Undecidable Neighbours}


In this subsection we show that the rest of the memory logics
presented in this work are undecidable. We start by analyzing the
decidability of two variations of $\cMLRKM$.

We first show that adding just one nominal $i$ to $\cMLRKM$ is
sufficient to turn into undecidability. We denote with $\cMLRKM+i$
this extension of $\cMLRKM$.
%
We next show that the satisfiability problem of the logic $\cMLRKM$
restricted to the class of models with $S=\emptyset$ (i.e.\ the
logic $\cMLRKME$) is also undecidable.

From the undecidability of $\cMLRKME$, we will conclude the
undecidability of $\cMLRK$ and $\cMLRKE$. Additionally, every time
we show undecidability, we also prove lack of finite model property.

%The last issue that we will discuss in this paper
%is the undecidability of the satisfiability problem
%for both $\cMLRK$ and $\cMLRKE$.  The proof is an adaptation
%of the proof of undecidability of $\hlogic$ presented in~\cite{BS95}.

%We first prove that both languages lack the finite model property~\cite{BRV01}.


\begin{thm}\label{thm:tlmi:inf}
$\cMLRKM + i$ lacks the finite model property.
% There is a formula
%$\textit{Inf} \in \cMLRKM + i$ such that $\mathcal{M},w \models
%\textit{Inf}$ implies that the domain of $\mathcal{M}$ is infinite.
\end{thm}

\begin{pf}
Consider the following formulas:
$$
\begin{array}{rl}
(\textit{Back}) & i \land \bbox{r}\lnot i  \land \ttup{r} \top \land \bbox{r}\ttup{r} i \\
%&$``The root is irreflexive, and in symmetric relation with all its (nonempty) successors.''$\\
(\textit{Empty}) & \bbox{r}\lnot \known \land \bbox{r}\bbox{r}(\lnot i \to \lnot \known)\\
%&$``One step and two steps successors are not known.''$\\
(\textit{Spy}) & \bbox{r}\bbox{r}( \lnot i \to \ttup{r} (i \land \ttup{r} ( \known \land \lnot \ttup{r} ( \known \land \lnot i))))\\
%&$``All successors are irreflexive.''$\\
%(\textit{Irr}) & \lnot \ttup{r} \ttup{r} ( \lnot i \land \known))\\
(\textit{Succ}) & \bbox{r}\ttup{r} \lnot i \\
(\textit{no-3cyc}) & \lnot \ttup{r} \ttup{r} ( \lnot \known \land \ttup{r} ( \lnot \known \land \ttup{r} ( \known \land \lnot i)))\\
(\textit{Tran}) & \bbox{r} \ttup{r} ( i \land \bbox{r} ( \lnot \known \to \ttup{r}( i \land \ttup{r} ( \known \land  \ttup{r}(\known \land \lnot i) ))))\\
\end{array}
$$
%
Let $\textit{Inf}$ be $\textit{Back} \land \textit{Empty} \land
\textit{Spy} \land \textit{Irr} \land \textit{Succ} \land
\textit{3cyc} \land \textit{Tran}$. Let $\model=\diam{W, R, V,
\emptyset}$. Let us see that if $\model, w \models \textit{Inf}$,
then $W$ is infinite.
%
Suppose $\model, w \models \textit{Inf}$. Let $B = \{b \in W \mid
wRb\}$. Because $\textit{Back}$ is satisfied, $w \not \in B$, $B
\not= \emptyset$ and for all $b \in B$, $bRw$. Note that
\textit{Empty} sais that the one and two-step horizon is not known,
and also this implies that every state in $B$ is irreflexive.
Because $\textit{Spy}$ is satisfied, if $a \not= w$ and $a$ is a
successor of an element of $B$ then $a$ is also an element of $B$.
As $\textit{Succ}$ is satisfied at $w$, every point in $B$ has a
successor distinct from $w$. As $\textit{no-3cyc}$ is satisfied,
there cannot be $3$ different elements in $B$ forming a cycle, and
this sentence together with $\textit{Tran}$ force $R$ to
transitively order $B$.
%
It follows that $B$ is an unbounded strict partial order, hence
infinite, and so is $W$.
\end{pf}


We now show that this logic is undecidable by encoding the
$\omega\times\omega$ tiling problem (see~\cite{BGG97}). Following
the idea in~\cite{BS95}, we construct a spy point over the relation
$s$ which has access to every tile. The relations $u$ and $r$
represent moving up and to the right, respectively, from one tile to
the other. We code each type of tile with a fixed propositional
symbol $t_i$. With this encoding we define for each tiling problem
$T$, a formula $\varphi^T$ such that the set of tiles $T$ tiles
$\omega\times\omega$ iff $\varphi^T$ has a model.


\begin{thm}\label{thm:tlmi:und}
The satisfiability problem for $\cMLRKM + i$ is undecidable.
\end{thm}
\begin{pf}
Let $T=\{T_1,\dots,T_n\}$ be a set of tile types. Given a tile type
$T_i$, $u(T_i)$, $r(T_i)$, $d(T_i)$, $l(T_i)$ will represent the
colors of the up, right, down and left edges of $T_i$ respectively.
Define:

\begin{displaymath}
\begin{array}{rl}
(\textit{Back}) & i \land \bbox{s}\lnot i \land \ttup{s} \top \land \bbox{s}\ttup{s}i \land \bbox{s}\bbox{s}i\\
(\textit{Empty}) & \bbox{s}\lnot \known \land \bbox{s}\bbox{s}(\lnot i \to \lnot \known) \land \bbox{s}[\dag]\lnot \known \quad \dag \in \{r,u\}\\
(\textit{Spy}) & \bbox{s}[\dag]\ttup{s}(i \land \ttup{s}(\known \land \lnot\diam{\dag}\known)) \quad \dag \in \{r,u\}\\
(\textit{Grid}) & \bbox{s}\diam{\dag} \top \quad \dag \in \{r,u\}\\
(\textit{Func}) & \bbox{s}[\dag]\ttup{s}\ttup{s}(\known \land \diam{\dag}\known \land [\dag]\known) \quad \dag \in \{r,u\}\\
(\textit{Conf}) & \bbox{s}\ttup{u}\ttup{r}\ttup{s}\ttup{s}(\known \land \lnot \ttup{r}\known \land \ttup{u}\known \land \ttup{r}( \lnot \known \land (\ttup{u}(\known \land \lnot \ttup{r}\known)))\\
(\textit{UR-no-Cycle}) & \bbox{s}\bbox{u}\bbox{r}\lnot \known \land \bbox{s}\bbox{r}\bbox{u}\lnot \known\\
(\textit{URU-no-Cycle}) & \bbox{s}\bbox{u}\bbox{r}\bbox{u}\lnot \known\\
(\textit{Unique}) & \bbox{s} \left( \bigvee_{1\leq i\leq n} t_i \wedge \bigwedge_{1\leq i < j \leq n } (t_i\to \lnot t_j)\right) \\
(\textit{Vert}) & \bbox{s} \bigwedge_{1\leq i\leq n} \left( t_i \to \ttup{u} \bigvee_{1\leq j\leq n,u(T_i)=d(T_j)}  t_j\right) \\
(\textit{Horiz}) & \bbox{s} \bigwedge_{1\leq i\leq n} \left( t_i \to \ttup{r} \bigvee_{1\leq j\leq n,r(T_i)=l(T_j)}  t_j\right) \\
\end{array}
\end{displaymath}
Let the formula $\varphi^T$ be the conjunction of all the above
formulas. We show that $T$ tiles $\omega\times\omega$ iff
$\varphi^T$ is satisfiable.

Suppose $\model,w\models\varphi^T$. Observe that (\textit{Back}) and
(\textit{Spy}), together with (\textit{Empty}) impose $w$ to be a
spy point over all its $S$-accessible states of $\model$ (and also
force $U$ and $R$ to be irreflexive and asymmetric). These
$S$-accessible states will be the tiles. From this it follows that
$[s]\psi$ holds at $w$ iff $\psi$ is true at every tile.
Additionally, $\diam{s}\diam{s}\psi$ holds at tile $v$ iff $\psi$ is
true at some tile (maybe the same one).

Taking the above points into account, one can establish the
following. (\textit{Grid}) states that from every tile there is
another tile moving up (that is, following the $U$-relation). The
same holds for the right direction (following the $R$-relation).
(\textit{Func}) forces that $U$ and $R$ are both functionals, given
that (\textit{Back}) and (\textit{Spy}) guarantee irreflexivity and
asymmetry of $U$ and $R$ respectively. (\textit{Conf}) imposes that
the tiles are arranged in a grid pattern. To make its job,
(\textit{Conf}) needs the composed relation $U\circ R$ to be
irreflexive, and to avoid a cycle with the $URU$ pattern. This job
is done by (\textit{UR-Irr}) and (\textit{URU-noCycle})
respectively.

All the formulas we discuss up to now configure the grid. The last
three ensure that every tile has a unique type $t_i$, and that the
colors of the tiles match properly. From this, it easily follows
that $\model$ is a tiling of $\omega\times\omega$.

For the converse, suppose $f:\omega\times\omega\to T$ is a tiling of
$\omega\times\omega$. We define the model
$\model=\diam{W,\{S,U,R\},V,\emptyset}$ as follows:
\begin{itemize}
\item $W=\omega\times\omega \cup \{w\}$
\item $S=\{(w,v),(v,w)\mid v\in\omega\times\omega\}$  (hence $w$ will act as the spy
point)
\item $U=\{((x,y),(x,y+1))\mid x,y\in\omega\}$
\item $R=\{((x,y),(x+1,y))\mid x,y\in\omega\}$
\item $V(p)=\{w\}$; $V(t_i)=\{x\mid x\in\omega\times\omega, f(x)=T_i\}$
\end{itemize}
The reader may verify that, by construction,
$\model,w\models\varphi^T$.
\end{pf}

We now turn to the case for $\cMLRKME$.


\begin{thm}\label{thm:tlme:inf}
$\cMLRKME$ lacks the finite model property.
%
%There is a formula $\textit{Inf} \in \cMLRKME$ such that
%$\mathcal{M},w \models \textit{Inf}$ implies that the domain of
%$\mathcal{M}$ is an infinite set.
\end{thm}

\begin{pf}
Consider the following formulas:
$$
\begin{array}{rl}
(\textit{Back}) & p \land \bbox{r} \lnot p  \land \ttup{r} \top \land \bbox{r}\ttup{r} (\known \land p) \\
%&$``The root is irreflexive, and in symmetric relation with all its (nonempty) successors.''$\\
(\textit{Spy}) & \bbox{r}\bbox{r}( \lnot p \to \ttup{r} (\known \land p \land \ttup{r} ( \known \land \lnot \ttup{r} ( \known \land \lnot p))))\\
%&$``A 2-step successor is a 1-step successor, with a non-symmetric relation between them.''$\\
(\textit{Irr}) & \lnot \ttup{r} ( \lnot p \land \ttup{r} ( \lnot p \land \known)))\\
%&$``All successors are irreflexive.''$\\
(\textit{Succ}) & \bbox{r}\ttup{r} \lnot p \\
%&$``All successors have a successor different from the root.''$\\
(\textit{no-3cyc}) & \lnot \ttup{r}   \ttup{r} ( \lnot \known \land \ttup{r} ( \lnot \known \land \ttup{r} ( \known \land \lnot p)))\\
%&$``There are no cycles of 3 elements between successors of the root.''$\\
(\textit{Tran}) & \bbox{r} ( \lnot p \to \ttup{r} ( \known \land p \land \bbox{r} ( \lnot p \land \lnot \known \to \ttup{r}( \known \land p \land \ttup{r} ( \known \land \lnot p \land  \ttup{r}(\known \land \lnot p) )))))\\
%&$``Every pair of successors $u$ and $v$ are related (either $uRv$ or $vRu$).''$\\
\end{array}
$$

Let $\textit{Inf}$ be $\textit{Back} \land \textit{Spy} \land
\textit{Irr} \land \textit{Succ} \land \textit{no-3cyc} \land
\textit{Tran}$. Let $\model=\diam{W, R, V, \emptyset}$. One can
follow the idea used in the proof of Theorem~\ref{thm:tlmi:inf} to
show that if $\model, w \models \textit{Inf}$, then $\model$ is
infinite.

In that proof, the use of $i$ was to identify the spy point. We now
don't have nominals, but using $S=\emptyset$, we can replace $i$
with a clever combination of a propositional symbol $p$ and
$\known$. Since we want to univocally identify the spy point and $p$
may hold in many states, sometimes we need to use $\known$ to
distinguish which one is the spy point (in the proof of
Theorem~\ref{thm:tlmi:inf}, this was not necessary because the
nominal $i$ is true at only one state).

Notice that \textit{Back}, \textit{Spy}, \textit{Succ}, and
\textit{no-3cyc} are very similar to the ones in the proof of
Theorem~\ref{thm:tlmi:inf}, \textit{Irr} forces $R$ to be
irreflexive and \textit{Tran} says that every pair of successors $u$
and $v$ are related (either $uRv$ or $vRu$), and this together with
the other formulas, implies that $R$ is transitive.

%
%Suppose $\model, w \models \textit{Inf}$. Let $B = \{b \in W \mid
%wRb\}$. Because $\textit{Back}$ is satisfied, $w \not \in B$, $B
%\not= \emptyset$ and for all $b \in B$, $bRw$. Because
%$\textit{Spy}$ is satisfied, if $a \not= w$ and $a$ is a successor
%of an element of $B$ then $a$ is also an element of $B$. As
%$\textit{Irr}$ is satisfied at $w$, every state in $B$ is
%irreflexive. As $\textit{Succ}$ is satisfied at $w$, every point in
%$B$ has a successor distinct from $w$. As $\textit{3cyc}$ is
%satisfied, there cannot be $3$ different elements in $B$ forming a
%cycle, and this sentence together with $\textit{Tran}$ force $R$ to
%transitively order $B$.
%
%It follows that $B$ is an unbounded strict partial order, hence
%infinite, and so is $W$.
\end{pf}


In the same way as before, we now show that the satisfiability
problem for $\cMLRKME$ is undecidable, by encoding the $\omega
\times \omega$ tiling problem.

\begin{thm}\label{thm:tlme:und}
The satisfiability problem for $\cMLRKME$ is undecidable.
\end{thm}
\begin{pf}
Let $T=\{T_1,\dots,T_n\}$ be a set of tile types. Given a tile type
$T_i$, $u(T_i)$, $r(T_i)$, $d(T_i)$, $l(T_i)$ will represent the
colors of the up, right, down and left edges of $T_i$ respectively.
Define:

\begin{displaymath}
\begin{array}{rl}
(\textit{Back}) & p \land \bbox{s}\lnot p \land \ttup{s} \top \land \bbox{s}\ttup{s}(\known \land p) \land \bbox{s}\bbox{s}(\known \land p)\\
(\textit{Spy}) & \bbox{s}[\dag](\lnot p \land \ttup{s}(\known \land p \land \ttup{s}(\known \land \lnot\diam{\dag}\known))) \quad \dag \in \{r,u\}\\
(\textit{Grid}) & \bbox{s}\diam{\dag} \top \quad \dag \in \{r,u\}\\
(\textit{Func}) & \bbox{s}[\dag]\ttup{s}\ttup{s}(\known \land \diam{\dag}\known \land [\dag]\known) \quad \dag \in \{r,u\}\\
(\textit{Conf}) & \bbox{s}\ttup{u}\ttup{r}\ttup{s}\ttup{s}(\known \land
\lnot \ttup{r}\known \land \ttup{u}\known \land
\ttup{r}\ttup{u}(\known \land \lnot \ttup{r}\known))\\
(\textit{URU-no-Cycle}) & \bbox{s}\bbox{u}\bbox{r}\lnot \known \land \bbox{s}\bbox{r}\bbox{u}\lnot \known\\
(\textit{URU-no-Cycle}) & \bbox{s}\bbox{u}\bbox{r}\bbox{u}\lnot \known\\
(\textit{Unique}) & \bbox{s} \left( \bigvee_{1\leq i\leq n} t_i \wedge \bigwedge_{1\leq i < j \leq n } (t_i\to \lnot t_j)\right) \\
(\textit{Vert}) & \bbox{s} \bigwedge_{1\leq i\leq n} \left( t_i \to \ttup{u} \bigvee_{1\leq j\leq n,u(T_i)=d(T_j)}  t_j\right) \\
(\textit{Horiz}) & \bbox{s} \bigwedge_{1\leq i\leq n} \left( t_i \to
\ttup{r} \bigvee_{1\leq j\leq n,r(T_i)=l(T_j)}  t_j\right)
\end{array}
\end{displaymath}
Let the formula $\varphi^T$ be the conjunction of all the above
formulas. One can show that $T$ tiles $\omega\times\omega$ iff
$\varphi^T$ is satisfiable, following the ideas used in the proof of
Theorem~\ref{thm:tlmi:und} and the observation in the proof of
Theorem~\ref{thm:tlme:inf}.

%Suppose $\model,w\models\varphi^T$. Observe that (\textit{Back}) and
%(\textit{Spy}) impose $w$ to be a spy point over all its
%$S$-accessible states of $\model$ (and also force $U$ and $R$ to be
%irreflexive and asymmetric). These $S$-accessible states will be the
%tiles. From this it follows that $[S]\psi$ holds at $w$ iff $\psi$
%is true at every tile. Additionally, $\diam{S}\diam{S}\psi$ holds at
%tile $v$ iff $\psi$ is true at some tile (maybe the same one).
%
%Taking the above points into account, one can establish the
%following. (\textit{Grid}) states that from every tile there is
%another tile moving up (that is, following the $U$-relation). The
%same holds for the right direction (following the $R$-relation).
%(\textit{Func}) forces that $U$ and $R$ are both functionals, given
%that (\textit{Back}) and (\textit{Spy}) guarantee irreflexivity and
%asymmetry of $U$ and $R$ respectively. (\textit{Conf}) imposes that
%the tiles are arranged in a grid pattern. To make its job,
%(\textit{Conf}) needs the composed relation $U\circ R$ to be
%irreflexive, and to avoid a cycle with the $URU$ pattern. This job
%is done by (\textit{UR-Irr}) and (\textit{URU-noCycle})
%respectively.
%
%All the formulas we discuss up to now configure the grid. The last
%three ensure that every tile has a unique type $t_i$, and that the
%colors of the tiles match properly. From this, it easily follows
%that $\model$ is a tiling of $\omega\times\omega$.
%
%For the converse, suppose $f:\omega\times\omega\to T$ is a tiling of
%$\omega\times\omega$. We define the model
%$\model=\diam{W,\{S,U,R\},V,\emptyset}$ as follows:
%\begin{itemize}
%\item $W=\omega\times\omega \cup \{w\}$
%\item $S=\{(w,v),(v,w)\mid v\in\omega\times\omega\}$  (hence $w$ will act as the spy
%point)
%\item $U=\{((x,y),(x,y+1))\mid x,y\in\omega\}$
%\item $R=\{((x,y),(x+1,y))\mid x,y\in\omega\}$
%\item $V(p)=\{w\}$; $V(t_i)=\{x\mid x\in\omega\times\omega, f(x)=T_i\}$
%\end{itemize}
%The reader may verify that, by construction,
%$\model,w\models\varphi^T$.
%
\end{pf}

\subsection{The Undecidable $\cMLRK$ and $\cMLRKE$}

\begin{thm}
$\cMLRKE$ lacks the finite model property and is undecidable.
\end{thm}

\begin{pf}
Straightforward from the effective translation given as a result of $\cMLRKME \leq \cMLRKE$
(Theorem~\ref{thm:four-le-two}) and theorems~\ref{thm:tlme:inf}
and~\ref{thm:tlme:und}.
\end{pf}


To prove failure of the finite model property for the case $\cMLRK$
we first notice that the following lemma is easy to establish (we only
state it for the mono-modal case; a similar result is true in the
multi-modal case).
Failure of the finite model property is then a direct consequence.

\begin{lem}\label{lem:boxes}
Let $\varphi$ be a formula of modal depth $d$. If
{\em $\diam{W,R_r,V,S},w\models$} \linebreak {\em $\left(\bigwedge_{i=0}^d [r]^i \lnot
\known\right) \wedge \varphi$} then $\diam{W,R_r,V,\emptyset},w\models
\varphi$.
\end{lem}

\begin{cor}
{\em $\cMLRK$} lacks the finite model property.
\end{cor}

\begin{pf}
Using Lemma~\ref{lem:boxes}, one can easily see that the formula
$\textit{Inf}\land \left(\bigwedge_{i=0}^4 [r]^i \lnot
\known\right)$, where $\textit{Inf}$ is the one in the proof of
Theorem~\ref{thm:tlme:inf}, forces an infinite model.
\end{pf}



\begin{cor}
The satisfiability problem for {\em $\cMLRK$} is undecidable.
\end{cor}
%
\begin{pf}
Using the idea of Lemma~\ref{lem:boxes} and the formula $\varphi^T$
in the proof of Theorem~\ref{thm:tlme:und}, we can obtain a formula
$\psi$ such that if $\model,w\models\psi$ then $\model$ is a tiling
of $\omega \times \omega$. For the converse, we can build exactly
the same model as in the proof of Theorem~\ref{thm:tlme:und} and
check that it satisfies $\psi$.
\end{pf}
